Measurement Bias from Address Aliasing

### 2016-01-09 paper.tex
Move microkernel result out of introduction
With shorter article, no need to tease this in the beginning.

%\pgfplotstableread{bin/microkernel-cycles-core2.dat}{\microkernelcoretwotable}
\pgfplotstableread{bin/microkernel-core.dat}{\tblmicrokernelcore}
\begin{figure}[t]
  \begin{tikzpicture}
    \begin{axis}[
        title = Core,
        width = \textwidth/2.1,
        height = 4.5cm,
        font = \footnotesize,
        xlabel=Bytes added to environment,
        ylabel=Cycles,
        domain = 0:4096,
        xtick = {0,1024,...,4096},
        xmin = 0,
        xmax = 4096,
        cycle list name = exotic
      ]
      \addplot[ycomb] table[x expr = \thisrowno{0}*16, y = cycles:u] \tblmicrokernelcore ;
    \end{axis}
  \end{tikzpicture}
  \caption{\label{fig:motivation}Performance counter data running the micro-kernel from Figure~\ref{fig:microkernel} under different environment sizes. Notice the clear spike at around 3200 bytes, where the number of cycles executed nearly doubles. Measuring performance in this configuration would give a biased result.}
\end{figure}

\begin{figure}[t]
  \begin{lstlisting}[frame=single, xleftmargin=.01\textwidth, xrightmargin=.01\textwidth]
static int i, j, k;
int main() {
    int g = 0, inc = 1;
    for (; g < 65536; g++) {
        i += inc;
        j += inc;
        k += inc;
    }
    return 0;
}
  \end{lstlisting}
  \caption{\label{fig:microkernel}Microkernel first presented by Mytkowicz et al.\cite{Mytkowicz:2009:WrongData}, showing bias to environment size.}
\end{figure}

@@

In Figure~\ref{fig:motivation}, we have reproduced a result originally presented in a paper by Mytkowicz et al.\cite{Mytkowicz:2009:WrongData}, showing how the number of cycles executed for a simple program (Figure~\ref{fig:microkernel}) depends on environment variables.
This example provides the basis for our analysis.
We show how some instances of measurement bias, such as the above, can be explained by \emph{address aliasing}, an artifact of a presumably Intel specific optimization on the out-of-order execution pipeline.
The CPU uses a heuristic for determining whether loads are dependent on previous stores, comparing only the 12 least significant virtual address bits.
Aliased memory accesses can produce false dependencies, and incur performance penalties.
We look at how address aliasing can trigger bias from two external factors; size of system environment variables and configuration of dynamic heap allocation libraries.

@@

@@

\pgfplotstableread{bin/microkernel-nehalem.dat}{\tblmicrokernelnehalem}
\pgfplotstableread{bin/microkernel-ivybridge.dat}{\tblmicrokernelivybridge}
\begin{figure}[t]
  \begin{tikzpicture}
    \begin{axis}[
        title = Nehalem (HT),
        y tick label style={
        /pgf/number format/.cd,
            fixed,
            fixed zerofill,
            precision=1,
        /tikz/.cd
        },
        width = \textwidth/2.1,
        height = 4.5cm,
        xlabel = Bytes added to environment,
        ylabel = Cycles,
        font = \footnotesize,
        domain = 0:4096,
        xtick = {0,1024,...,4096},
        xmin = 0,
        xmax = 4096,
        cycle list name = exotic
      ]
      \addplot[ycomb] table[x expr = \thisrowno{0}*16, y = cycles:u] \tblmicrokernelnehalem ;
    \end{axis}
  \end{tikzpicture}
  \begin{tikzpicture}
    \begin{axis}[
        title = Ivy Bridge,
        y tick label style={
        /pgf/number format/.cd,
            fixed,
            fixed zerofill,
            precision=1,
        /tikz/.cd
        },
        width = \textwidth/2.1,
        height = 4.5cm,
        xlabel = Bytes added to environment,
        ylabel = Cycles,
        font = \footnotesize,
        domain = 0:4096,
        xtick = {0,1024,...,4096},
        xmin = 0,
        xmax = 4096,
        cycle list name = exotic
      ]
      \addplot[ycomb] table[x expr = \thisrowno{0}*16, y = cycles:u] \tblmicrokernelivybridge ;
    \end{axis}
  \end{tikzpicture}
  \caption{\label{fig:microkernel-all-archs}Microkernel experiment on the ``Nehalem'' and ``Ivy Bridge'' microarchitectures.}
\end{figure}

\subsection{Other Architectures}
While we show only the analysis for ``Haswell'' here, the same approach gave similar results for all the CPUs we tested.
Figure~\ref{fig:microkernel-all-archs} shows additional cycle count plots for ``Nehalem'' and ``Ivy Bridge'', revealing the same alias behavior.\footnote{For these results we also pinned the process to a single core during measurements using \texttt{taskset}, which helps remove some of the random noise.}
The experiment was also done for the older ``Core'' setup, shown in Figure~\ref{fig:motivation}.
On this architecture, the \emph{load\_blocks.overlap\_store} event is used to indicate address aliasing.
Notably, this event was also found to be significant in the original paper, where the authors used a similar setup~\cite{Mytkowicz:2009:WrongData}.
We were able to verify that partial address overlap was indeed occuring at the same time this event is spiking, and draw the conclusion that address aliasing must be the underlying explanation.
Still, this only explains part of what is going on in Figure~\ref{fig:motivation}, as there seem to be yet another bias effect at every fourth environment increment.
These smaller spikes correlate with the performance event \emph{load\_blocks.std}, which is described as ``loads blocked by a preceding store with unknown data''~\cite[Table 19-17]{Volume3B}.
%Our assumption is that this is a different phenomenon, not related to 4K aliasing.
We were only able to observe the effect on this particular architecture, and did not investigate the root cause.% is related to 4K aliasing.

>>>>

@@

%In Figure~\ref{fig:motivation}, we have reproduced a result originally presented in a paper by Mytkowicz et al.\cite{Mytkowicz:2009:WrongData}, showing how the number of cycles executed for a simple program (Figure~\ref{fig:microkernel}) depends on environment variables.
%This example provides the basis for our analysis.
We show how some instances of measurement bias can be explained by \emph{address aliasing}, an artifact of a presumably Intel specific optimization on the out-of-order execution pipeline.
The CPU uses a heuristic for determining whether loads are dependent on previous stores, comparing only the 12 least significant virtual address bits.
Aliased memory accesses can produce false dependencies, and incur performance penalties.
We look at how address aliasing can trigger bias from two external factors; size of system environment variables and configuration of dynamic heap allocation libraries.

@@

% Original code presented my Mytkowicz et al.
\begin{figure}[t]
  \begin{lstlisting}[frame=single, xleftmargin=.01\textwidth, xrightmargin=.01\textwidth]
static int i, j, k;
int main() {
    int g = 0, inc = 1;
    for (; g < 65536; g++) {
        i += inc;
        j += inc;
        k += inc;
    }
    return 0;
}
  \end{lstlisting}
  \caption{\label{fig:microkernel}Microkernel first presented by Mytkowicz et al.\cite{Mytkowicz:2009:WrongData}, showing bias to environment size.}
\end{figure}

@@

%\pgfplotstableread{bin/microkernel-cycles-core2.dat}{\microkernelcoretwotable}
\pgfplotstableread{bin/microkernel-core.dat}{\tblmicrokernelcore}
\pgfplotstableread{bin/microkernel-nehalem.dat}{\tblmicrokernelnehalem}
\pgfplotstableread{bin/microkernel-ivybridge.dat}{\tblmicrokernelivybridge}
\begin{figure}[t]
  \begin{tikzpicture}
    \begin{axis}[
        title = Core,
        width = \textwidth/2.1,
        height = 4.5cm,
        font = \footnotesize,
        xlabel=Bytes added to environment,
        ylabel=Cycles,
        domain = 0:4096,
        xtick = {0,1024,...,4096},
        xmin = 0,
        xmax = 4096,
        cycle list name = exotic
      ]
      \addplot[ycomb] table[x expr = \thisrowno{0}*16, y = cycles:u] \tblmicrokernelcore ;
    \end{axis}
  \end{tikzpicture}
  %\caption{\label{fig:motivation}Performance counter data running the micro-kernel from Figure~\ref{fig:microkernel} under different environment sizes. Notice the clear spike at around 3200 bytes, where the number of cycles executed nearly doubles. Measuring performance in this configuration would give a biased result.}
  \begin{tikzpicture}
    \begin{axis}[
        title = Nehalem (HT),
        y tick label style={
        /pgf/number format/.cd,
            fixed,
            fixed zerofill,
            precision=1,
        /tikz/.cd
        },
        width = \textwidth/2.1,
        height = 4.5cm,
        xlabel = Bytes added to environment,
        ylabel = Cycles,
        font = \footnotesize,
        domain = 0:4096,
        xtick = {0,1024,...,4096},
        xmin = 0,
        xmax = 4096,
        cycle list name = exotic
      ]
      \addplot[ycomb] table[x expr = \thisrowno{0}*16, y = cycles:u] \tblmicrokernelnehalem ;
    \end{axis}
  \end{tikzpicture}
  \begin{tikzpicture}
    \begin{axis}[
        title = Ivy Bridge,
        y tick label style={
        /pgf/number format/.cd,
            fixed,
            fixed zerofill,
            precision=1,
        /tikz/.cd
        },
        width = \textwidth/2.1,
        height = 4.5cm,
        xlabel = Bytes added to environment,
        ylabel = Cycles,
        font = \footnotesize,
        domain = 0:4096,
        xtick = {0,1024,...,4096},
        xmin = 0,
        xmax = 4096,
        cycle list name = exotic
      ]
      \addplot[ycomb] table[x expr = \thisrowno{0}*16, y = cycles:u] \tblmicrokernelivybridge ;
    \end{axis}
  \end{tikzpicture}
  \caption{\label{fig:microkernel-all-archs}Microkernel experiment on the ``Nehalem'' and ``Ivy Bridge'' microarchitectures.}
\end{figure}

\subsection{Other Architectures}
While we show only the analysis for ``Haswell'' here, the same approach gave similar results for all the CPUs we tested.
Figure~\ref{fig:microkernel-all-archs} shows additional cycle count plots for ``Core'', ``Nehalem'' and ``Ivy Bridge'', revealing the same alias behavior.%\footnote{For these results we also pinned the process to a single core during measurements using \texttt{taskset}, which helps remove some of the random noise.}
%The experiment was also done for the older ``Core'' setup, shown in Figure~\ref{fig:motivation}.
On the older ``Core'' architecture, \emph{load\_blocks.overlap\_store} is used to indicate address aliasing.
Notably, this event was also found to be significant in the original paper, where the authors used a similar setup~\cite{Mytkowicz:2009:WrongData}.
We were able to verify that partial address overlap was indeed occuring at the same time this event is spiking, and draw the conclusion that address aliasing must be the underlying explanation.
Still, this only explains part of what is going on in Figure~\ref{fig:motivation}, as there seem to be yet another bias effect at every fourth environment increment.
These smaller spikes correlate with the performance event \emph{load\_blocks.std}, which is described as ``loads blocked by a preceding store with unknown data''~\cite[Table 19-17]{Volume3B}.
%Our assumption is that this is a different phenomenon, not related to 4K aliasing.
We were only able to observe the effect on this particular architecture, and did not investigate the root cause.% is related to 4K aliasing.
